Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2016-02-08 paper.tex
modded paper

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray X-ray luminosity was limited to 10$^{43}$~erg~s$^{-1}$. 
Here, we introduce a simple treatment of clumping, and find that a filling factor of $\sim0.01$ 
moderates the ionization state sufficently for BAL features to form in the 
rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with AGN X-ray properties and the wind produces strong line emission in, e.g., \la\ and \civline\ at low inclinations. At high inclinations, the spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all emission lines seen in quasar spectra with the correct equivalent-width ratios, and we find an angular dependence of emission-line equivalent width despite the similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that biconical winds can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}),
which are reprocessed by `macro-atoms' with associated transition probabilities.
We use the Sobolev approximation \citep[e.g.][]{sobolev1957,sobolev1960,rybickihummer1978}
to compute the location and optical depth of line interactions.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as macro-atoms, because the process is computationally
intensive, and we expect recombination to be particularly important
in determining their level populations and resultant line emission.
We are also especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-ions. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and
prepares for a more accurate treatment of other physical processes in future.

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$, in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-ions. This dilute approximation 
is not required for macro-atom levels.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
Adiabatic cooling in the cell is calculated using the estimator
\begin{equation}
C_{A} = f_V k_B T_e V (\nabla \cdot v) \left(n_e + \sum\limits_{i=1}^{ions} N_i\right),
\end{equation}
where $V$ is the cell volume, $T_e$ is the electron temperature,
$n_e$ is the electron density, $N_j$ is the number density of ion $j$
and $\nabla \cdot v$ is the divergence of the 
velocity field at the centre of the cell. The sum is over all ions 
included in the simulation.
For consistency, $k$-packets then have a finite 
destruction probability of 
\begin{equation}
P_{k,A} = \frac{C_{A}}{C_{tot}},
\end{equation}
where $C_{tot}$ is the sum over all cooling rates in the cell.

\subsection{Atomic Data}

We use the same atomic data  described by LK02 as updated by H13 and M15, 
with the addition of direct (collisional) ionization and recombination
data from \cite{dere2007}. 
Photoionization cross-sections are from \textsc{Topbase} \citep{cunto1993} and \cite{vfky}.
%Dielectronic and 
Radiative recombination rate coefficients are taken from 
the \textsc{Chianti} database version 7.0 \citep{dere1997,landi2012}.
We use ground state recombination rates from \cite{badnell2006} where available,
and otherwise default to calculating recombination rates from the Milne
relation. Free-free Gaunt factors are from \cite{sutherland1998}.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping the wind with a volume filling factor of $0.01$ 
moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where the continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

>>>>

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray
luminosity was limited to 10$^{43}$~erg~s$^{-1}$. 
Here, we introduce a simple treatment of clumping, and find that a filling factor of $\sim0.01$ 
moderates the ionization state sufficently for BAL features to form in the 
rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with AGN X-ray properties and the wind produces strong line emission in, e.g., \la\ and \civline\ at low inclinations. At high inclinations, the spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all emission lines seen in quasar spectra with the correct equivalent-width ratios, and we find an angular dependence of emission-line equivalent width despite the similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that biconical winds can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}),
which are reprocessed by `macro-atoms' with associated transition probabilities.
We use the Sobolev approximation \citep[e.g.][]{sobolev1957,sobolev1960,rybickihummer1978}
to compute the location and optical depth of line interactions.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple-atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as macro-atoms, because the process is computationally
intensive, and we expect recombination to be particularly important
in determining their level populations and resultant line emission.
We are also especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

Macro-atoms have their ion and level populations derived from
MC rate estimators as described by Lucy (2002,2003). 
Previously (LK02, H13, M15), we used a modified Saha approach to 
calculate the ionization fractions
of simple-atoms. As part of  this effort, we have 
now improved {\sc PYTHON} to explicitly solve the 
rate equations between ions in non-LTE. This dispenses with a number of small assumptions 
made in the modified Saha approach, is more numerically stable, and
prepares for a more accurate treatment of other physical processes in future.

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$, in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-atoms. This dilute approximation 
is not required for macro-atom levels.

@@

We include  free-free, bound-free and bound-bound heating
and cooling processes in the model. 
For radiative transfer purposes
we treat electron scattering in the Thomson limit, 
but take full account of Compton heating and cooling when
calculating the thermal balance of the plasma (see H13).
Adiabatic cooling in a cell is calculated using the estimator
\begin{equation}
C_{A} = f_V k_B T_e V (\nabla \cdot v) \left(n_e + \sum\limits_{i=1}^{ions} N_i\right),
\end{equation}
where $V$ is the cell volume, $T_e$ is the electron temperature,
$n_e$ is the electron density, $N_i$ is the number density of ion $i$
and $\nabla \cdot v$ is the divergence of the 
velocity field at the centre of the cell. The sum is over all ions 
included in the simulation.
For consistency, $k$-packets then have a destruction probability of 
\begin{equation}
P_{k,A} = \frac{C_{A}}{C_{tot}},
\end{equation}
where $C_{tot}$ is the sum over all cooling rates in the cell.

\subsection{Atomic Data}
We use the same atomic data  described by LK02 as updated by H13 and M15.
As in M15, macro-atom level information comes from \textsc{Topbase} \citep{cunto1993}.
H and He~\textsc{ii} are treated with 10 levels, and we include one He~\textsc{i} 
level. This is because He is almost always completely ionized 
and so He~\textsc{i} is unimportant in the simulation.
% with the addition of direct (collisional) ionization and recombination
% data from \cite{dere2007}. 
Photoionization cross-sections are from \textsc{Topbase} \citep{cunto1993} and \cite{vfky}.
Dielectronic and Radiative recombination rate coefficients for simple-atoms are taken from 
the \textsc{Chianti} database version 7.0 \citep{dere1997,landi2012}.
We use ground state recombination rates from \cite{badnell2006} where available,
and otherwise default to calculating recombination rates from the Milne
relation. Free-free Gaunt factors are from \cite{sutherland1998}.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping the wind with a volume filling factor of $0.01$ 
moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show a \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where the continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.
